% ============================================================
% Chapter III — Methodology
% ============================================================
\section{Methodology}

\subsection{Research Design}

This study employs a quantitative, cross-sectional research design using
multivariate statistical methods. The primary inferential tool is
\textbf{multinomial logistic regression} estimated via maximum likelihood
(MLE), with \textbf{leave-one-group-out cross-validation} (LOGO-CV) serving
as the machine learning validation framework. The research follows a
predictive modeling approach: features are engineered from within-race
telemetry data, the statistical model is fitted using MLE, and
generalizability is assessed through cross-validation across circuits.

\subsection{Data Source}

All data were obtained from the \textbf{FastF1} Python API, which provides
programmatic access to official F1 timing, telemetry, and session data. Three
races from the 2023 Formula One World Championship were selected:

\begin{enumerate}
  \item \textbf{Bahrain Grand Prix} (Sakhir International Circuit) ---
    permanent desert circuit with high tire degradation.
  \item \textbf{Saudi Arabian Grand Prix} (Jeddah Corniche Circuit) ---
    high-speed street circuit with minimal run-off.
  \item \textbf{Australian Grand Prix} (Albert Park Circuit) ---
    parkland circuit with mixed slow-speed and medium-speed corners.
\end{enumerate}

These three circuits were chosen to provide architectural diversity:
permanent, street, and parkland layouts ensure that the model encounters
meaningfully different track characteristics rather than memorizing a single
circuit profile.

\subsection{Participants and Sampling}

The unit of analysis is the \textbf{individual lap}. Every lap from every
driver across all three events was included in the dataset, yielding a total
of \textbf{2,880 observations} after data cleaning. No sampling was performed;
the dataset represents the complete population of completed laps from the three
selected races.

\subsection{Instrumentation}

The sole data collection instrument is the FastF1 Python library (v3.8.2),
which interfaces with the official F1 data feed. The following data fields
were extracted:

\begin{itemize}
  \item \texttt{LapTime}, \texttt{LapNumber} --- lap timing
  \item \texttt{TyreLife} --- cumulative laps on current tire set (tire age)
  \item \texttt{Compound} --- tire compound (Soft, Medium, Hard)
  \item \texttt{Position} --- running position at lap completion
  \item \texttt{GridPosition} --- starting grid position
  \item \texttt{Driver} --- driver abbreviation
  \item \texttt{Status} --- finishing status (Finished, Retired, etc.)
  \item \texttt{ClassifiedPosition} --- official race classification
\end{itemize}

\subsection{Data Collection Procedure}

Data collection proceeded in five stages:

\begin{enumerate}[label=Phase \arabic*:, leftmargin=*]
  \item \textbf{Session Loading.} Each race session was loaded via
    \texttt{fastf1.get\_session()} with telemetry, lap, and weather data
    enabled. The FastF1 cache system was used to store downloaded data
    locally.

  \item \textbf{Multi-Race Stacking.} All laps from the three sessions were
    concatenated into a single dataframe. Each observation was tagged with a
    \texttt{Race\_ID} (e.g., \texttt{2023\_Bahrain}) to enable group-aware
    cross-validation.

  \item \textbf{Target Variable Definition.} The \texttt{Finishing\_Tier}
    variable was mapped from \texttt{Position}:
    \begin{itemize}
      \item \textbf{Podium:} Position $\leq 3$
      \item \textbf{Point Scorer:} $4 \leq$ Position $\leq 10$
      \item \textbf{No Points:} Position $> 10$
    \end{itemize}

  \item \textbf{DNF Protocol.} Drivers with \texttt{Status = "Retired"} or
    \texttt{ClassifiedPosition = "R"} were flagged, and all their completed
    laps were reassigned to the ``No Points'' tier. This prevents a driver
    running P2 pace before a mechanical failure from contaminating the Podium
    cluster.

  \item \textbf{Feature Engineering.} Three metric predictors were constructed:
    \begin{itemize}
      \item \textit{Fuel Mass Proxy:} A linear burn-off model starting at
        110\,kg and decreasing proportionally with lap count:
        \[
          \text{Fuel Mass Proxy} = 110 - \frac{110}{\text{Max Lap}} \times
          \text{Lap Number}
        \]
      \item \textit{Tire Age:} Directly sourced from \texttt{TyreLife}.
      \item \textit{Degradation $\times$ Fuel:} The mean-centered interaction:
        \[
          \text{Deg} \times \text{Fuel} =
          (\text{Tire Age} - \bar{x}_{\text{Tire Age}}) \times
          (\text{Fuel Mass} - \bar{x}_{\text{Fuel Mass}})
        \]
      \item \textit{Tire Compound Dummies:} $k{-}1$ dummy coding with the
        first alphabetical compound (Hard) as baseline, yielding
        \texttt{Tire\_SOFT} and \texttt{Tire\_MEDIUM}.
    \end{itemize}
\end{enumerate}

\subsection{Statistical Treatment}

The following multivariate methods were applied:

\subsubsection{Box's M Test}

Box's M test evaluates the null hypothesis that the variance--covariance
matrices of the continuous predictors are equal across the three finishing
tiers. This test served as a method selection gate: if the assumption of
homogeneous covariance matrices were met, Multiple Discriminant Analysis (MDA)
would be appropriate; rejection of this assumption motivated the choice of
multinomial logistic regression instead.

\subsubsection{Variance Inflation Factor (VIF)}

VIF was computed for all predictors to detect multicollinearity. The standard
threshold of VIF $\geq 10$ was used to flag problematic collinearity. This was
particularly important for validating that the mean-centered interaction term
and its component variables remain structurally independent.

\subsubsection{Multinomial Logistic Regression (MLE)}

The multinomial logistic regression model estimates simultaneous log-odds
equations for each non-baseline tier:

\[
  \ln\!\left(\frac{P(Y = k)}{P(Y = \text{No Points})}\right) =
  \beta_{0k} + \beta_{1k} \cdot \text{Grid} +
  \beta_{2k} \cdot \text{TireAge} +
  \beta_{3k} \cdot \text{Fuel} +
  \beta_{4k} \cdot \text{Deg}\times\text{Fuel} +
  \beta_{5k} \cdot \text{Tire\_MED} +
  \beta_{6k} \cdot \text{Tire\_SOFT}
\]

where $k \in \{\text{Podium}, \text{Point Scorer}\}$ and ``No Points'' is the
reference category. The model was fitted using the \texttt{multinom()} function
from the \texttt{nnet} R package.

\subsubsection{Statistical Inference}

Z-scores and exact $p$-values were computed from the MLE coefficient estimates
and standard errors. The significance threshold was set at $\alpha = 0.05$
with $p < 0.01$ noted for highly significant predictors.

\subsubsection{Machine Learning Validation}

A scikit-learn \texttt{LogisticRegression} model with \texttt{class\_weight=
"balanced"} was validated using LOGO-CV across the three races. In each fold,
the model was trained on two races and tested on the held-out third race.
Balanced class weighting was applied to mitigate the natural imbalance where
``No Points'' laps outnumber ``Podium'' laps. Performance was evaluated using
precision, recall, F1-score, and overall accuracy.
